---
author: Nelson-Gon

---

# missplore: Missing Data Explorer

As the name suggests, `missplore` is intended to make missing data exploration as smooth and as easy as possible. While several packages exist to handle the same, differences in syntax and preferences will often lead to different people using different packages.

`missplore` therefore hopes to fill this gap by offering a very simple yet robust way to explore missingness. 

# Installation

We can install `missplore` as follows:

```{r}
# install.packages("devtools")
devtools::install_github("Nelson-Gon/missplore")

```

# Currently available functions.

1. `get_na_counts`

This provides a convenient way to show the number of missing values columnwise. It is relatively fast(tests done on about 400,000 rows, took a few microseconds.)

```{r}
get_na_counts(airquality)
```

The above might be less useful if one would like to get the results by group. In that case, one can set `grouped` to `TRUE` and provide a vector of names in `grouping_cols` that will be used for grouping. 

```{r}

test <- structure(list(Subject = structure(c(1L, 1L, 2L, 2L), .Label = c("A", 
"B"), class = "factor"), res = c(NA, 1, 2, 3), ID = structure(c(1L, 
1L, 2L, 2L), .Label = c("1", "2"), class = "factor")), class = "data.frame", row.names = c(NA, 
-4L))

get_na_counts(test, grouped = TRUE, grouping_cols = "ID")

```



2. `percent_missing`

This is a very simple to use but quick way to take a look at the percentage of data that is missing columnwise.

```{r}

percent_missing(airquality)

```

I don't want percentages, give me decimals


```{r}

percent_missing(airquality, percent = FALSE)

```

Again, we can get the results by group by setting `grouped` to `TRUE`

```{r}
percent_missing(test, grouped = TRUE, grouping_cols = "Subject")

```

3. `recode_as_na`

As the name might imply, this converts any value or vector of values with `NA` i.e we take a value such as "missing" and convert it to R's known handler for missing values(`NA`).

To use the function out of the box(with default arguments), one simply does something like:

```{r}

dummy_test <- data.frame(ID = c("A","B","B","A"), 
                         values = c("n/a",NA,"Yes","No"))
# Convert n/a to NA
recode_as_na(dummy_test, value = "n/a")

```

Great, but I want to do so for specific columns not the entire dataset. You can do this by setting `subset_df` to `TRUE` and providing column names to `subset_cols`. 

```{r}

another_dummy <- data.frame(ID = 1:5, Subject = 7:11, 
Change = c("missing","n/a",2:4 ))
# Only change values at the column Change
recode_as_na(another_dummy, subset_df = TRUE,
             subset_cols = "Change", value = c("n/a",
                                               "missing"))
                                               

```


